---
title: "BCB 724 Homework 2"
author: "Matthew Sutcliffe"
date: '`r Sys.Date()`'
output: pdf_document
geometry: "a4paper, margin=1cm"
---

Load libraries and data
```{r, results='hide', message=FALSE}
library(caret)
library(glmnet)
library(pROC)

set.seed(0)

metadata <- read.csv(file = "poore-2020-metadata-subset.csv", stringsAsFactors = TRUE)[, -1]
```

One-hot encode the 6 feature columns
```{r}
binary_features <- model.matrix(object = formula( ~ . - 1), data = metadata[, -1])

dim(binary_features)
```

Combine with target "Ovarian Serous Cystadenocarcinoma"
```{r}
x <- cbind(data.frame(
    "OSC" = as.numeric(metadata$disease_type == "Ovarian Serous Cystadenocarcinoma")
    ),binary_features)
```

50/50 train/test split
```{r}
trainIndex <- createDataPartition(y = x$OSC, p = 0.5, list = FALSE, times = 1)

x_train <- x[ trainIndex, ]
x_test  <- x[-trainIndex, ]
```

# Model Fitting

Fit a multivariate logistic regression model with LASSO penalty.
```{r}
fit <- cv.glmnet(x = as.matrix(x_train[, -1]),
                 y = as.matrix(x_train[,  1, drop = FALSE]),
                 family = "binomial", alpha = 1)
plot(fit)
```

What features does it select?
```{r}
selected_features <- as.matrix(predict(object = fit, s = "lambda.min", type = "coefficients"))[, 1]
selected_features <- selected_features[selected_features != 0] |> sort(decreasing = TRUE)

head(selected_features)
```
There are 54 features selected. Interestingly, one of the top features include "tissue_source_site_label = Gynecologic Oncology Group", which seems highly relevant for predicting ovarian cancer.


Let's look at the predicted probabilities for the training set and testing set
```{r, fig.width=8, fig.height=5}
prediction_train <- predict(object = fit,
                            newx = as.matrix(x_train[, -1]),
                            s = "lambda.min",
                            type = "response")[, 1]
prediction_test  <- predict(object = fit,
                            newx = as.matrix(x_test[, -1]),
                            s = "lambda.min",
                            type = "response")[, 1]

par(mfrow = c(1, 2))
hist(prediction_train, las = 1)
hist(prediction_test, las = 1)
```

All the probabilities are very close to 0 or 1, so regardless of whether the model is correct, it seems very confident in the predictions.

Calculate ROC curves
```{r, results='hide', message=FALSE}
x_train$prediction <- prediction_train
x_test$prediction <- prediction_test

roc_train <- roc(OSC ~ prediction, data = x_train)
roc_test <- roc(OSC ~ prediction, data = x_test)

auc_train <- auc(roc_train) |> as.numeric()
auc_test <- auc(roc_test) |> as.numeric()
```
# Results

```{r, echo=FALSE, fig.width=8, fig.height=5}
par(mfrow = c(1, 2), pty="s")
plot(x = NA, y = NA,
     xlim = c(0, 1), ylim = c(0, 1),
     xlab = "1 - Specificity", ylab = "Sensitivity",
     axes = FALSE, xaxs = "i", yaxs = "i")
axis(side = 1)
axis(side = 2, las = 1)
par(xpd = FALSE)
abline(a = 0, b = 1, col = "gray")
par(xpd = TRUE)
plot_train_x <- 1 - roc_train$specificities
plot_train_y <- roc_train$sensitivities
lines(x = plot_train_x, y = plot_train_y,
      lwd = 4)
plot_test_x <- 1 - roc_test$specificities
plot_test_y <- roc_test$sensitivities
lines(x = plot_test_x, y = plot_test_y,
      lwd = 4, col = "red")
legend("bottomright", legend = c("train", "test"), border = NA, lty = c(1, 1), col = c("black", "red"), lwd = 4)
xmax <- 0.2
ymin <- 0.8
rect(0, ymin, xmax, 1, lty = "dotted")

plot(x = NA, y = NA,
     xlim = c(0, xmax), ylim = c(ymin, 1),
     xlab = "1 - Specificity", ylab = "Sensitivity",
     axes = FALSE, xaxs = "i", main = "zoom in")
axis(side = 1)
axis(side = 2, las = 1)
par(xpd = TRUE)

plot_train_x <- 1 - roc_train$specificities
plot_train_y <- roc_train$sensitivities
lines(x = plot_train_x[plot_train_x < xmax & plot_train_y > ymin],
      y = plot_train_y[plot_train_x < xmax & plot_train_y > ymin],
      lwd = 4)
plot_test_x <- 1 - roc_test$specificities
plot_test_y <- roc_test$sensitivities
lines(x = plot_test_x[plot_test_x < xmax & plot_test_y > ymin],
      y = plot_test_y[plot_test_x < xmax & plot_test_y > ymin],
      lwd = 4, col = "red")
text(x = 0.07, y = 0.9, paste("AUC_train =", round(auc_train, 3),"\n"), col = "black", adj = c(0, 0))
text(x = 0.07, y = 0.9, paste("AUC_test =", round(auc_test, 3),"\n"), col = "red", adj = c(0, 1))
```